\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-11-01 paper.tex
add group dependance on association vector Psi

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\sA}, \] with $\sA = \sum_{a=1}^\cA \imath_a$ and so that
\begin{equation}
  \label{eq:def-asso-features}
  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
\end{equation}
where $\Psi(t) \in \R^{L\sA}$ denotes the association features vector at time $t$.

\paragraph{Specification of the functionals $(\varphi_a)_{a\in \cA}$.}

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_a(t, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at each uncensored observed event times as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, b_i^l) \Big\}.\]

@@

\begin{proof}
Fix any $\beta \in \R^p$ and denote 
\begin{equation*}
    k_\beta(x) =  \frac{1}{2}\norm{ x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)}_2^2 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
\end{equation*}
for all $x \in \R^p$. Given that $\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big)_j=\max(0, 1-\lambda_1/|\beta_j|)\beta_j$~\citep{bach2011optimization}, $k_\beta$ is clearly strictly convex. Denoting $x^\star$ its unique minimizer,~\eqref{eq:prox-def} entails 
\[\text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big) = \argmin_{x \in \R^p} k_\beta(x) = \{x^\star\}.\]
Then, it follows that
\begin{equation}
    \label{eq:opt_gl}
    0 \in \partial k(x^\star) = x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial\big( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l}\big)
\end{equation}
where $\partial k$ denotes the subgradient~\citep{rockafellar1974} of $k$.
Now denoting
\[h_\beta(x) = \frac{1}{2}\norm{x - \beta}_2^2 + \lambda_1 \norm{x}_1 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l} \] 
for all $x \in \R^p$ such that
\[\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}(\beta) = \argmin_{x\in \R^p} h_\beta(x),\]
it remains to be proved that $0 \in \partial h_\beta(x^\star)$.

@@

Let us introduce the following useful notations
\begin{equation*}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
    \end{array}
  \right. 
\end{equation*}
We also denote 
\[ \cI_1 = 
\begin{bmatrix}
  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
\end{bmatrix}
\in \R^{n \times J}
\quad \text{and} \quad 
\cI_2 = 
\begin{bmatrix}
  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
\end{bmatrix}
\in \R^{n \times J} \]
some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance, the quantities
\begin{align*}
f(t_i, \delta_i| \breve{b}, G_i = k ; \theta^{(w)}) = &\big(\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k^{(w)}}(\breve{b}) \big)^{\delta_i} \\ 
&\times \exp \Big\{ - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}) \ind{\{t_j^u \leq t_i\}} \Big\}
\end{align*}
for $i = 1, \ldots, n$ and $\breve{b} \in S^{(w)}$ can be represented in the following matrix 
\[ \underset{J}{\text{sum}} \big( \boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_1 \big)^{\boldsymbol{\delta}} \exp \big\{- \underset{J}{\text{sum}} \big(\boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_2 \big) \big\} \in \R^{n \times 2N} \]
where 
\begin{equation*}
  \left\{
    \begin{array}{ll}
    \boldsymbol{g^4_{\gamma_k}} \in \R^{2N \times J} \text{ with } \big[\boldsymbol{g^4_{\gamma_k}}\big]_{i, j} = \tilde g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}_i) , \\
    \boldsymbol{\lambda_0^{(w)}} = \big(\lambda_0^{(w)}(t_1^u), \ldots, \lambda_0^{(w)}(t_J^u) \big)^\top \in \R^J, \\
    \boldsymbol{\delta} = (\delta_1, \ldots, \delta_n)^\top \in \R^n.
    \end{array}
  \right.
\end{equation*}
Another example is the update of the baseline hazard~\eqref{eq:update_lambda_0} as
\[\boldsymbol{\lambda_0^{(w+1)}} = \underset{n}{\text{sum}}( \boldsymbol{\delta} \cI_1) / \underset{n}{\text{sum }} \underset{k}{\text{sum }} (\boldsymbol{g^4_{\gamma_k}} \cI_2 \boldsymbol{\hat \pi^{(w)}})\]
with $\boldsymbol{\hat \pi^{(w)}} \in \R^{2 \times n} \text{ and }  \left\{
    \begin{array}{ll}
    \boldsymbol{\hat \pi_{1,\bullet}^{(w)}}= (\hat \pi_{11}^{\theta^{(w)}}, \ldots, \hat \pi_{n1}^{\theta^{(w)}})^\top \in \R^n, \\
    \boldsymbol{\hat\pi_{0,\bullet}^{(w)}} = 1 - \boldsymbol{\hat\pi_{1,\bullet}^{(w)}}.
    \end{array}
  \right.$

>>>>

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_{k,a}(t, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_{k,a}$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_{k,a})$.
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\sA}, \] with $\sA = \sum_{a=1}^\cA \imath_a$ and so that
\begin{equation}
  \label{eq:def-asso-features}
  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi_k(t) \big\},
\end{equation}
where $\Psi_k(t) \in \R^{L\sA}$ denotes the association features vector at time $t$.

\paragraph{Specification of the functionals $(\varphi_{k,a})_{a\in \cA}$.}

The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework. In spite of that, rationale for selecting shared associations has received little attention. We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
\begin{table}[htb]
\centering
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_{k,a}(t, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}
\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
\label{table:shared_associations}
\end{table}

@@

We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
It can be written
\begin{align*}
  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_{k,i}(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_{k,i}(s) \big\} \dd s \\
  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
\end{align*}
Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_{k,i}(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at each uncensored observed event times as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_{k,i}(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

@@

In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_{k,a}(t_i^{max}, b_i^l) \Big\}.\]

@@

\begin{proof}
Fix any $\beta \in \R^p$ and denote 
\begin{equation*}
    k_\beta(x) =  \frac{1}{2}\norm{ x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)}_2^2 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
\end{equation*}
for all $x \in \R^p$. Given that $\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big)_j=\max(0, 1-\lambda_1/|\beta_j|)\beta_j$~\citep{bach2011optimization}, $k_\beta$ is clearly strictly convex. Denoting $x^\star$ its unique minimizer,~\eqref{eq:prox-def} entails 
\[\text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big) = \argmin_{x \in \R^p} k_\beta(x) = \{x^\star\}.\]
Then, it follows that
\begin{equation}
    \label{eq:opt_gl}
    0 \in \partial k_\beta(x^\star) = x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial\big( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l}\big)
\end{equation}
where $\partial k$ denotes the subgradient~\citep{rockafellar1974} of $k$.
Now denoting
\[h_\beta(x) = \frac{1}{2}\norm{x - \beta}_2^2 + \lambda_1 \norm{x}_1 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l} \] 
for all $x \in \R^p$ such that
\[\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}(\beta) = \argmin_{x\in \R^p} h_\beta(x),\]
it remains to be proved that $0 \in \partial h_\beta(x^\star)$.

@@

Let us introduce the following useful notations
\begin{equation*}
  \left\{
    \begin{array}{ll}
    \tilde g^1&: b_i \mapsto b_i,\\
    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
    \tilde g^3_{s, k}&: b_i \mapsto \Psi_{k,i}(s), \\
    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s, k}(b_i) \big\}, \\
    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s, k}(b_i)^\top g^4_{s, k}(b_i).
    \end{array}
  \right. 
\end{equation*}
We also denote 
\[ \cI_1 = 
\begin{bmatrix}
  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
\end{bmatrix}
\in \R^{n \times J}
\quad \text{and} \quad 
\cI_2 = 
\begin{bmatrix}
  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
  \vdots &  \ddots & \vdots \\
  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
\end{bmatrix}
\in \R^{n \times J} \]
some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance, the quantities
\begin{align*}
f(t_i, \delta_i| \breve{b}, G_i = k ; \theta^{(w)}) = &\big(\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k^{(w)}}(\breve{b}) \big)^{\delta_i} \\ 
&\times \exp \Big\{ - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}) \ind{\{t_j^u \leq t_i\}} \Big\}
\end{align*}
for $i = 1, \ldots, n$ and $\breve{b} \in S^{(w)}$ can be represented in the following matrix 
\[ \underset{J}{\text{sum}} \big( \boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_1 \big)^{\boldsymbol{\delta}} \exp \big\{- \underset{J}{\text{sum}} \big(\boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_2 \big) \big\} \in \R^{n \times 2N} \]
where 
\begin{equation*}
  \left\{
    \begin{array}{ll}
    \boldsymbol{g^4_{\gamma_k}} \in \R^{2N \times J} \text{ with } \big[\boldsymbol{g^4_{\gamma_k}}\big]_{i, j} = \tilde g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}_i) , \\
    \boldsymbol{\lambda_0^{(w)}} = \big(\lambda_0^{(w)}(t_1^u), \ldots, \lambda_0^{(w)}(t_J^u) \big)^\top \in \R^J, \\
    \boldsymbol{\delta} = (\delta_1, \ldots, \delta_n)^\top \in \R^n.
    \end{array}
  \right.
\end{equation*}
Another example is the update of the baseline hazard~\eqref{eq:update_lambda_0} as
\[\boldsymbol{\lambda_0^{(w+1)}} = \underset{n}{\text{sum}}( \boldsymbol{\delta} \cI_1) / \underset{n}{\text{sum }} \underset{k}{\text{sum }} (\boldsymbol{g^4_{\gamma_k}} \cI_2 \boldsymbol{\hat \pi^{(w)}})\]
with $\boldsymbol{\hat \pi^{(w)}} \in \R^{2 \times n} \text{ and }  \left\{
    \begin{array}{ll}
    \boldsymbol{\hat \pi_{1,\bullet}^{(w)}}= (\hat \pi_{11}^{\theta^{(w)}}, \ldots, \hat \pi_{n1}^{\theta^{(w)}})^\top \in \R^n, \\
    \boldsymbol{\hat\pi_{0,\bullet}^{(w)}} = 1 - \boldsymbol{\hat\pi_{1,\bullet}^{(w)}}.
    \end{array}
  \right.$
